Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-06-10 paper/ms.tex
Add mock HST images
Also some other clean up

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/fig1.pdf}
\caption{\textbf{Projected surface densities of simulated barred galaxies with
and without the interstellar medium.} The upper panels show an \Nbody{} only
simulation while the lower panels show a simulation which includes the SMUGGLE
model for the ISM. Columns show different points in time, separated by about
$1\,\textrm{Gyr}$. We can see that in the \Nbody{} run, the bar grows in
length and strength. In the SMUGGLE run, the bar remains at approximately the
same length and strength over the course of the simulation. The \Nbody{} model
is identical to the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} Stellar disks in isolation are prone to developing bars,
\cite{1971ApJ...168..343H} but the presence of a strong spherical potential
(e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk
against bar formation.\cite{1973ApJ...186..467O, 1976AJ.....81...30H} Many
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on a bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in a dark matter halo in a manner
akin to dynamical friction. This wake lags a bar and thus exerts a negative
torque on it.

@@

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one such example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628.\cite{2009AA...499L..25C} This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.

@@

\item [Acknowledgements] We would like to thank Greg~L. Bryan, Neal~J. Evans,
Drummond~B. Fielding, Keith Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson,
Kathryn~V. Johnston, Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro,
Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful discussions.
A.B. would like to thank Todd Phillips for helpful discussions. Some
computations in this paper were run on the FASRC Cannon cluster supported by
the FAS Division of Science Research Computing Group at Harvard University.
Resources supporting this work were also provided by the NASA High-End
Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
Division at Ames Research Center. A.B. was supported by the Future
Investigators in NASA Earth and Space Science and Technology (FINESST) award
number 80NSSC20K1536 during the completion of this work. L.H. funding. E.D.
funding. J.Q. acknowledges support from NSF grant
AST-2008490. L.V.S. funding. P.T. acknowledges support from NSF grant
AST-1909933, AST-2008490, and NASA ATP Grant 80NSSC20K0502. M.V. acknowledges
support through NASA ATP 19-ATP19-0019, 19-ATP19-0020, 19-ATP19-0167, and NSF
grants AST-1814053, AST-1814259, AST-1909831, AST-2007355 and AST-2107724.

@@

\noindent
{\bf Initial Setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
\texttt{MakeNewDisk}.\cite{2005MNRAS.361..776S} The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile.\cite{1990ApJ...356..359H} All \Nbody{} runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas
disk which is radially exponential and in vertical gravito-hydrodynamic
balance. We modified the radial profile of this code in order to allow us to
generate a disk with a constant surface density within some cut-off radius,
and then exponentially declining beyond that radius with the scale-length of
the stellar disk. We used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$.

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS \Nbody{} disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the \Nbody{} run to be the time at which we added the
gas phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the \Nbody{} disk when the bar is forming). We made one
additional modification when stitching the gas disk together with the \Nbody{}
disk - we created a hole within the central $4\,\textrm{kpc}$. This hole
guards against an initial dramatic infall of gas within the bar region, which
we found to destroy the bar. Models of the Milky Way's gas surface density
profile indicate a reduction in gas in the bar region from $\sim1$ to
$\sim4\,\textrm{kpc}$,\cite{2022ApJ...929L..18E} and so our practice of
allowing the gas distribution to have a hole in the central region is
consistent with our choice to begin the simulations with a bar already formed.

@@

\noindent
{\bf SMUGGLE Model}
\\
\noindent
We use the Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE) model
\cite{2019MNRAS.489.4233M} implemented within the moving-mesh, finite-volume
hydrodynamics and gravity code AREPO \cite{2010MNRAS.401..791S}. The SMUGGLE model
additionally includes radiative heating and cooling, star
formation, and stellar feedback. Explicit gas cooling and heating of the
multi-phase interstellar medium is implemented, covering temperature ranges
between $10$ and $10^8\,\textrm{K}$.

@@

\noindent
{\bf Smooth Interstellar Medium Model}
\\
\noindent
In addition to the SMUGGLE model, we considered a simpler model of the
interstellar medium based upon ref.\cite{2003MNRAS.339..289S} In this model,
the multiphase nature of the interstellar medium is modeled in a subgrid
manner by allowing each resolution element to have a ``cold'' and ``hot''
component, with the equation of state of the gas suitably modified. Gas is
allowed to interchange between the ``cold'' and ``hot'' components through
processes such as cooling and stellar feedback. Cold gas is allowed to undergo
star formation. The pattern speed evolution of this setup is given in Extended
Data Fig.~\ref{fig:GFM}, which shows that the pattern speed evolves in a
similar manner to the SMUGGLE case shown in Fig.~\ref{fig:prop}.

@@

\noindent
{\bf Varying Gas Fraction}
\\
\noindent
We also performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Extended Data Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{kpc}^2$ slows down at a similar rate to the \Nbody{} case.

@@

After the disk has been decomposed into a trapped and untrapped component, we
measured the bar length as being the radius $R_b$ which encapsulates $99\%$ of
the stars identified as being trapped in the bar, allowing for some outliers.
For the computation of the torque, we used the tree algorithm in
\texttt{MakeNewDisk}\cite{2005MNRAS.361..776S} customized to be accessible
from \texttt{Python} using \texttt{Cython}. This algorithm is based on the
\texttt{TREESPH} code.\cite{1989ApJS...70..419H} We constructed a tree using
only the star particles identified as being trapped in the bar using an
opening angle of $0.35$. We then queried the tree at the locations of all
resolution elements in the other components and computed the torque of the bar
on such components. The torque on the bar by the other components is simply
the negative of the torque on the other components by the bar. A similar
method was applied in measuring the torque for the disk when the pattern speed
is kept constant, as is done in Fig.~\ref{fig:equil}.

@@

\noindent
{\bf Plotting Details}
\\
\noindent
We saved snapshots in intervals of $0.005$ in the time units of the
simulation, $\textrm{kpc}/(\textrm{km}/\textrm{s})$, which is very nearly
equal to $1\,\textrm{Gyr}$ (it is $\sim0.977\,\textrm{Gyr}$). Therefore,
throughout this work we referred to the native code time unit as
$\textrm{Gyr}$. None of our results are sensitive to this choice.

>>>>

\DeclareMathOperator\arcsinh{arcsinh}

@@

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/fig1.pdf}
\caption{\textbf{Synthetic Hubble Space Telescope images of our simulations with
and without the interstellar medium.} The upper panels show an \Nbody{} only
simulation while the lower panels show a simulation which includes the SMUGGLE
model for the interstellar medium. Each panel is $20\,\textrm{kpc}$ to a side.
Columns show different points in time, separated by about $1\,\textrm{Gyr}$.
We can see that in the \Nbody{} run, the bar grows in length and strength. In
the SMUGGLE run, the bar remains at approximately the same length and strength
over the course of the simulation. The \Nbody{} model is identical to the
GALAKOS model, discussed in the text. Details on the production of these
synthetic images is given in the Methods section.}\label{fig:overview}
\end{figure*}

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} Stellar disks in isolation are prone to developing bars,
\cite{1971ApJ...168..343H} but the presence of a strong spherical potential
(e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk
against bar formation.\cite{1973ApJ...186..467O, 1976AJ.....81...30H} Many
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on a bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in a dark matter halo in a manner
akin to dynamical friction. This wake lags a bar and thus exerts a negative
torque on it. For a review of bar dynamics, see ref.\cite{1993RPPh...56..173S}

@@

We found that below a certain gas fraction, bars should still be able to slow
down. Therefore, we expect barred spiral galaxies which have been gas-poor for
extended periods of time to be rotating very slowly. We therefore predict that
observations which target such galaxies (e.g., lenticular barred
galaxies\cite{2009ARAA..47..159B}) would find slowly rotating bars. There does
exist one example of a galaxy which is known to be a slow rotator -- the
low surface brightness galaxy UGC 628.\cite{2009AA...499L..25C} This galaxy
has been studied in detail by ref.\cite{2016MNRAS.463.1751C}, who note that it
indeed has a low gas fraction for galaxies of its type. We predict a general
trend that bars in gas-rich spiral galaxies should rotate quickly while bars
in gas-poor spiral galaxies should rotate slowly.

@@

\item [Acknowledgements] We would like to thank Greg~L. Bryan, Neal~J. Evans,
Drummond~B. Fielding, Keith Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson,
Kathryn~V. Johnston, Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro,
Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful discussions.
A.B. would like to thank Todd Phillips for helpful discussions. Some
computations in this paper were run on the FASRC Cannon cluster supported by
the FAS Division of Science Research Computing Group at Harvard University.
Resources supporting this work were also provided by the NASA High-End
Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
Division at Ames Research Center. This research has made use of the Spanish
Virtual Observatory (https://svo.cab.inta-csic.es) project funded by
MCIN/AEI/10.13039/501100011033/ through grant PID2020-112949GB-I00. A.B. was
supported by the Future Investigators in NASA Earth and Space Science and
Technology (FINESST) award number 80NSSC20K1536 during the completion of this
work. L.H. funding. E.D. funding. C.C. funding. J.Q. acknowledges support from
NSF grant AST-2008490. L.V.S. is grateful for financial support from NASA ATP
80NSSC20K0566, NSF AST 1817233 and NSF CAREER 1945310 grants. P.T.
acknowledges support from NSF grant AST-1909933, AST-2008490, and NASA ATP
Grant 80NSSC20K0502. M.V. acknowledges support through NASA ATP 19-ATP19-0019,
19-ATP19-0020, 19-ATP19-0167, and NSF grants AST-1814053, AST-1814259,
AST-1909831, AST-2007355 and AST-2107724.

@@

\noindent
{\bf Initial setup}
\\
\noindent
The initial setup of the galactic disk used in this work follows closely the
GALAKOS model\cite{2020ApJ...890..117D}, which uses a modified version of
\texttt{MakeNewDisk}.\cite{2005MNRAS.361..776S} The GALAKOS model has three
components - a radially exponential and vertically isothermal stellar disk,
and a stellar bulge and dark matter halo following a Hernquist
profile.\cite{1990ApJ...356..359H} All \Nbody{} runs in this work used the same
setup parameters as the GALAKOS disk, more details of which can be found in
the original paper.

The addition of the gas phase was done as follows. The version of
\texttt{MakeNewDisk} used for the original GALAKOS model can generate a gas
disk which is radially exponential and in vertical gravito-hydrodynamic
balance. We modified the radial profile of this code in order to allow us to
generate a disk with a constant surface density within some cut-off radius,
and then exponentially declining beyond that radius with the scale-length of
the stellar disk. We used an initial surface density of
$20\,M_{\odot}/\textrm{pc}^2$ and a cut-off radius of $9.3\,\textrm{kpc}$. The
initial gas disk is generated with a temperature of $10^4\,\textrm{K}$ and
solar metallicity.

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS \Nbody{} disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the \Nbody{} run to be the time at which we added the
gas phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the \Nbody{} disk when the bar is forming). We made one
additional modification when stitching the gas disk together with the \Nbody{}
disk - we created a hole within the central $4\,\textrm{kpc}$. This hole
guards against an initial dramatic infall of gas within the bar region, which
we found to destroy the bar. It is not uncommon for observed barrred galaxies
to have gas deficits in the bar region (though not in the very
center),\cite{1993RPPh...56..173S} and so our practice of allowing the gas
distribution to have a hole in the central region is consistent with our
choice to begin the simulations with a bar already formed.

@@

\noindent
{\bf SMUGGLE model}
\\
\noindent
We use the Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE) model
\cite{2019MNRAS.489.4233M} implemented within the moving-mesh, finite-volume
hydrodynamics and gravity code AREPO \cite{2010MNRAS.401..791S}. The SMUGGLE model
additionally includes radiative heating and cooling, star
formation, and stellar feedback. Explicit gas cooling and heating of the
multi-phase interstellar medium is implemented, covering temperature ranges
between $10$ and $10^8\,\textrm{K}$.

@@

\noindent
{\bf Smooth interstellar medium model}
\\
\noindent
In addition to the SMUGGLE model, we considered a simpler model of the
interstellar medium based upon ref.\cite{2003MNRAS.339..289S} In this model,
the multiphase nature of the interstellar medium is modeled in a subgrid
manner by allowing each resolution element to have a ``cold'' and ``hot''
component, with the equation of state of the gas suitably modified. Gas is
allowed to interchange between the ``cold'' and ``hot'' components through
processes such as cooling and stellar feedback. Cold gas is allowed to undergo
star formation. The pattern speed evolution of this setup is given in Extended
Data Fig.~\ref{fig:GFM}, which shows that the pattern speed evolves in a
similar manner to the SMUGGLE case shown in Fig.~\ref{fig:prop}.

@@

\noindent
{\bf Varying gas fraction}
\\
\noindent
We also performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Extended Data Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{kpc}^2$ slows down at a similar rate to the \Nbody{} case.

@@

After the disk has been decomposed into a trapped and untrapped component, we
measured the bar length as being the radius $R_b$ which encapsulates $99\%$ of
the stars identified as being trapped in the bar, allowing for some outliers.
For the computation of the torque, we used the tree algorithm in
\texttt{MakeNewDisk}\cite{2005MNRAS.361..776S} customized to be accessible
from \texttt{Python} using \texttt{Cython}. This algorithm is based on the
\texttt{TREESPH} code.\cite{1989ApJS...70..419H} We constructed a tree with an opening angle of $0.35$ using
only the star particles identified as being trapped in the bar. We then queried the tree at the locations of all
resolution elements in the other components and computed the torque of the bar
on such components. The torque on the bar by the other components is simply
the negative of the torque on the other components by the bar. A similar
method was applied in measuring the torque for the disk when the pattern speed
is kept constant, as is done in Fig.~\ref{fig:equil}.

@@

\noindent
{\bf Idealized mock light images}
\\
\noindent
We created synthetic mock light images at $1024^2$ resolution in
Fig.~\ref{fig:overview} for the \textit{Hubble Space Telescope} using a
post-processed Monte Carlo radiation transfer simulation with the SKIRT9 code
including secondary emission from dust.\cite{2020AC....3100381C} Emission from
star particles was calculated assuming a Bruzual-Charlot spectral energy
distribution\cite{2003MNRAS.344.1000B} and a Chabrier initial stellar mass
function\cite{2003PASP..115..763C}. The luminosity from each star particle was
smoothed using a cubic spline kernel with each star particle being assigned a
smoothing length equal to its distance to its $32$nd closest neighbor. We
launched $10^9$ photon packets per simulation segment. For star particles
which were present in the initial simulation (i.e., initial disk and bulge
particles), we assumed that they have an age of $5\,\textrm{Gyr}$ and solar
metallicity. For star particles in the SMUGGLE simulation which are newly
formed, we used their recorded ages and metallicities.

For the SMUGGLE simulations, we included absorption and emission from dust. We
assumed the THEMIS (The Heterogeneous dust Evolution Model for Interstellar
Solids) dust model\cite{2017AA...602A..46J} to convert gas metallicity and
mass density to dust mass density. For the \Nbody{} simulations we ran the
SKIRT9 code in a mode without an obscuring medium, but otherwise applied the
same procedure.

We used the profiles for filters F814W, F606W, and F475W on the Advanced
Camera for Surveys (ACS) instrument for red, green, and blue pixel values,
respectively. Filtered fluxes were converted to RGB values using an $\arcsinh$
scaling following ref.\cite{2004PASP..116..133L} Scalings and cutoff values
were adjusted manually to create visually appealing images. Filter profiles
were provided by the Spanish Virtual Observatory Filter
Service.\cite{2012ivoa.rept.1015R, 2020sea..confE.182R}

\noindent
{\bf Plotting details}
\\
\noindent
We saved snapshots in intervals of $0.005$ in the time units of the
simulation, $\textrm{kpc}/(\textrm{km}/\textrm{s})$, which is very nearly
equal to $1\,\textrm{Gyr}$ (it is $\sim0.977\,\textrm{Gyr}$). Therefore,
throughout this work we referred to the native code time unit as
$\textrm{Gyr}$. None of our results are sensitive to this choice.
